---
title: "LIFE748 - ML and data analysis"
author: "Francesco Del Carratore"
format: html
editor: visual
---

GAI Declaration:

I, \[201945294\], declare that i have not employed ChatGPT in the creation of this script to help structure and refine R and shell scripts used in the analysis, debug code and resolve errors during implementation. All outputs generated evaluated, tested, and modified where necessary to ensure correctness and suitability for this project.

## Introduction

In this tutorial, we will analyze a simulated RNA-seq dataset. Our goal is to identify Differentially Expressed Genes (DEGs) between two conditions (Condition A and Condition B), visualize the results, and use both unsupervised and supervised machine learning techniques to classify the samples based on their gene expression profiles.

## Step 0: Load the required libraries

```{r warning=FALSE, message=FALSE }
library(DESeq2) # Bioconductor
library(ggplot2) # CRAN
library(EnhancedVolcano) # Bioconductor
library(pheatmap) # Bioconductor
library(caret) # CRAN
library(MASS) # CRAN
library(e1071) # CRAN
library(dplyr)
library(tidyr)
```

If these packages are not already available, you will need to install them either from CRAN (install.packages('package_name')), or from Bioconductor (BiocManager::install('package_name')

## Step 1: Load the Count Matrix

First, we load the raw count matrix and set up the sample metadata (phenotype data). We assume the first 25 columns belong to Condition A, and the next 25 belong to Condition B.

```{r}
counts <- read.csv("dataset_LIFE748_1.csv", row.names = 1)

colData <- data.frame(
  row.names = colnames(counts),
  Condition = c(rep('CondA',25),rep('CondB',25))
)

```

## Step 2: Normalize and Identify DEGs with DESeq2

We will use `DESeq2` to normalize the library sizes and perform statistical testing to identify which genes are differentially expressed between Condition B and Condition A.

```{r}
# 1. Create the DESeq2 dataset object
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = colData,
                              design = ~ Condition)

# 2. Run the DESeq2 pipeline
dds <- DESeq(dds)

# 3. Extract the results (Condition B vs Condition A)
res <- results(dds)

# 4. Convert this object into a dataframe for easier visualisation (optional)

res <- as.data.frame(res)

# 5. Extract normalized counts for downstream machine learning
vsd <- vst(dds, blind = FALSE) # Variance stabilizing transformation
normalized_counts <- assay(vsd)

```

## Step 3: Report DEGs (Table and Volcano Plot)

We define statistically significant DEGs as those with an adjusted p-value (FDR) \< 0.05 and an absolute log2 Fold Change \> 1.

```{r}
# Significant DEGs
sig_res <- res[!is.na(res$padj) & res$padj < 0.05 & abs(res$log2FoldChange) > 1, ]
sig_res <- sig_res[order(sig_res$padj), ]

# Convert to data frame
sig_res_df <- as.data.frame(sig_res)
sig_res_df$Gene <- rownames(sig_res_df)

# Summary counts
total_deg <- nrow(sig_res_df)
up_deg <- sum(sig_res_df$log2FoldChange > 1)
down_deg <- sum(sig_res_df$log2FoldChange < -1)

cat("Total DEGs:", total_deg, "\n")
cat("Upregulated DEGs:", up_deg, "\n")
cat("Downregulated DEGs:", down_deg, "\n")

# View top DEGs
head(sig_res_df[, c("Gene", "log2FoldChange", "padj")], 20)
```

Generate a volcano plot to visualise the results

```{r}
res_df <- as.data.frame(res)
res_df$Gene <- rownames(res_df)
res_df$Significant <- "Not significant"
res_df$Significant[!is.na(res_df$padj) & res_df$padj < 0.05 & abs(res_df$log2FoldChange) > 1] <- "Significant"

ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = Significant), alpha = 0.7) +
  scale_color_manual(values = c("grey70", "red")) +
  labs(
    title = "Volcano plot of differential expression",
    x = "log2 Fold Change",
    y = "-log10 adjusted p-value"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

```

## Step 4: Isolate Differentially Expressed Genes

To reduce noise and dimensionality for our machine learning algorithms, we will subset our normalized count matrix to only include the statistically significant DEGs identified above.

In order to make sure we don't lose too much information, let's relax a bit the thresholds for significance

```{r}
# Relaxed threshold for ML
ml_res <- res[!is.na(res$padj) & res$padj < 0.1 & abs(res$log2FoldChange) > 0.5, ]
ml_res <- ml_res[order(ml_res$padj), ]

ml_genes <- rownames(ml_res)

# Subset normalized counts
deg_matrix <- normalized_counts[ml_genes, , drop = FALSE]

# Samples as rows, genes as columns
ml_data <- as.data.frame(t(deg_matrix))
ml_data$Condition <- colData$Condition

cat("Genes retained for ML:", length(ml_genes), "\n")
dim(ml_data)
head(ml_data[, 1:min(6, ncol(ml_data))])
```

## Step 5: Unsupervised Clustering

Let's see if the samples naturally group by their condition without explicitly telling the algorithm which sample belongs to which group. We will use Hierarchical Clustering and K-means.

## k-means

```{r}
set.seed(123)

kmeans_input <- scale(ml_data[, !names(ml_data) %in% "Condition"])
km_res <- kmeans(kmeans_input, centers = 2, nstart = 25)

table(Cluster = km_res$cluster, Condition = ml_data$Condition)
```

## hierarchical clustering

```{r}
dist_mat <- dist(kmeans_input)
hc_res <- hclust(dist_mat, method = "complete")

plot(
  hc_res,
  labels = ml_data$Condition,
  main = "Hierarchical clustering of samples",
  xlab = "",
  sub = ""
)
```

Visualise the results (HINT: an heatmap would be perfect for this).

```{r}
annotation_col <- data.frame(Condition = ml_data$Condition)
rownames(annotation_col) <- rownames(ml_data)

pheatmap(
  deg_matrix,
  scale = "row",
  annotation_col = annotation_col,
  show_rownames = FALSE,
  main = "Heatmap of selected DEGs"
)
```

## Step 6: Supervised Learning (LDA and SVM)

Next, we train classifiers to predict the condition of a sample based on its DEG expression profile. We will split the data into a training set (70%) and a testing set (30%).

```{r}
set.seed(123)

ml_data$Condition <- factor(ml_data$Condition, levels = c("CondA", "CondB"))

trainIndex <- createDataPartition(ml_data$Condition, p = 0.7, list = FALSE)

trainingData <- ml_data[trainIndex, , drop = FALSE]
testingData  <- ml_data[-trainIndex, , drop = FALSE]

trainingData$Condition <- factor(trainingData$Condition)
testingData$Condition  <- factor(testingData$Condition)

dim(trainingData)
dim(testingData)
table(trainingData$Condition)
table(testingData$Condition)

# Logistic regression
glm_model <- glm(Condition ~ ., data = trainingData, family = binomial)

glm_probs <- predict(glm_model, newdata = testingData, type = "response")

levels(testingData$Condition)

positive_class <- levels(testingData$Condition)[2]
negative_class <- levels(testingData$Condition)[1]

glm_pred <- ifelse(glm_probs > 0.5, positive_class, negative_class)
glm_pred <- factor(glm_pred, levels = levels(testingData$Condition))
```

Train LDA (Linear Discriminant Analysis)

```{r}
lda_model <- lda(Condition ~ ., data = trainingData)

lda_pred <- predict(lda_model, newdata = testingData)$class
```

Train SVM (Support Vector Machine)

```{r}
svm_model <- svm(Condition ~ ., data = trainingData, kernel = "linear")

svm_pred <- predict(svm_model, newdata = testingData)
```

## Step 7: Compare Classifier Performance

Finally, we use confusion matrices to evaluate and compare how well LDA and SVM performed on the unseen test dataset.

```{r}
glm_cm <- confusionMatrix(glm_pred, testingData$Condition, positive = positive_class)
lda_cm <- confusionMatrix(lda_pred, testingData$Condition, positive = positive_class)
svm_cm <- confusionMatrix(svm_pred, testingData$Condition, positive = positive_class)

glm_cm
lda_cm
svm_cm

```

```{r}

perf_table <- data.frame(
  Model = c("Logistic regression", "LDA", "SVM"),
  Accuracy = c(
    unname(glm_cm$overall["Accuracy"]),
    unname(lda_cm$overall["Accuracy"]),
    unname(svm_cm$overall["Accuracy"])
  ),
  Sensitivity = c(
    unname(glm_cm$byClass["Sensitivity"]),
    unname(lda_cm$byClass["Sensitivity"]),
    unname(svm_cm$byClass["Sensitivity"])
  ),
  Specificity = c(
    unname(glm_cm$byClass["Specificity"]),
    unname(lda_cm$byClass["Specificity"]),
    unname(svm_cm$byClass["Specificity"])
  )
)

perf_table

perf_long <- perf_table %>%
  pivot_longer(cols = c(Accuracy, Sensitivity, Specificity),
               names_to = "Metric", values_to = "Value")

ggplot(perf_long, aes(x = Model, y = Value, fill = Metric)) +
  geom_col(position = "dodge") +
  ylim(0, 1) +
  labs(
    title = "Classifier performance comparison",
    x = "Model",
    y = "Performance"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 20, hjust = 1)
  )
```

## Step 8: Is this enough?

is this enough to compare the classifiers? What if you use a k-fold cross-validation approach? Will you get similar results with different random splits?

``` {#k-fold-cross-validation}
```

```{r}
set.seed(123)

cv_ctrl <- trainControl(
  method = "repeatedcv",
  number = 5,
  repeats = 5
)

cv_glm <- train(
  Condition ~ .,
  data = ml_data,
  method = "glm",
  family = "binomial",
  trControl = cv_ctrl
)

cv_lda <- train(
  Condition ~ .,
  data = ml_data,
  method = "lda",
  trControl = cv_ctrl
)

cv_svm <- train(
  Condition ~ .,
  data = ml_data,
  method = "svmLinear",
  trControl = cv_ctrl
)

cv_results <- data.frame(
  Model = c("Logistic regression", "LDA", "SVM"),
  CV_Accuracy = c(
    max(cv_glm$results$Accuracy),
    max(cv_lda$results$Accuracy),
    max(cv_svm$results$Accuracy)
  )
)

cv_results
```

```{r}
ggplot(cv_results, aes(x = Model, y = CV_Accuracy)) +
  geom_col() +
  ylim(0, 1) +
  labs(
    title = "Repeated cross-validation accuracy",
    x = "Model",
    y = "Cross-validated accuracy"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 20, hjust = 1)
  )
```

```{r}
top_markers <- sig_res_df[order(sig_res_df$padj), c("Gene", "log2FoldChange", "padj")]
head(top_markers, 20)
```
